Science Drivers for LSST Alert Distribution Latency

### 2020-02-27 draft-latency.tex
added draft-latency

% To add a short-form title:
% \title[Short title]{Title}
\title[Alerts Menu]{LSST Alerts: Science-Driven Options for Packets and their Distribution}

@@

\author{%
M.~L.~Graham and the Data Management System Science Team, with contributions from the Transients and Variable Stars Science Collaboration
}

@@

\setDocAbstract{%
\textcolor{red}{\bf EXTREME DRAFT STATE. DO NOT CITE. DO NOT READ. LOOK AWAY.}\\
A review and discussion of the variety of options for alert packets and their distribution, such as latency timescale, packet contents, pre-stream filtering, and so forth.
}

% Change history defined here.
% Order: oldest first.
% Fields: VERSION, DATE, DESCRIPTION, OWNER NAME.
% See LPM-51 for version number policy.
\setDocChangeRecord{%
  \addtohist{0}{2019-11-14}{Conception.}{Melissa Graham}
  \addtohist{0.1}{2020-01-09}{First draft ready for input from others.}{Melissa Graham}
}

@@

The purpose of this document is to consider the science motivation for, and impact of, various options for alert packet contents and their distribution timescale.
The contents and delivery of alert packets will play a major role in the scientific impact of LSST, in part because they are the only LSST data product that is both world public, as they are exempt from any proprietary period, and publicly accessible because they are distributed to community brokers, and not solely available via the LSST Science Platform (LSP).
In comparison, although the contents of the Prompt products database (PPDB) from which alerts are generated are also world public, access to the PPDB is restricted to individuals with data rights \citedsp{LDO-013} who have authenticated accounts in the LSP.

It is a requirement\reqparam{numStreams}\dmreq{0391} that the DMS be capable of supporting the transmission of at least 5 full alert streams within 60 seconds of image readout.
This is based in part on estimates of the alert stream data rate and the bandwidth allocated to alert distribution in the LSST data facility.
Reducing the packet size or increasing (slightly) the latency could enable the stream to be delivered to more brokers, and thereby increase the amount of alerts-related science done by the community.
The alert packets, as described in \citeds{LSE-163}, were designed to include all the relevant LSST data about a source that a broker might need to assess, classify, and prioritize the alert for follow-up observations \emph{within minutes} (i.e., on order the timescale of the 60 second alert distribution latency).
Reducing the contents should only be done if that information is easily and quickly obtainable elsewhere.

{\bf Document Overview ---} \S~\ref{sec:packets} discusses how removing stamps and histories from alerts could be scientifically beneficial for some brokers, as long as stamps are made publicly available with similar latency, and proposes additional alert elements to better identify potential transient host galaxies.
\S~\ref{sec:latency} find that most alerts-related science goals would be satisfied with a $\sim$5 minute alert timescale, but that the quickly evolving fields of fast radio bursts and gravitational wave events may soon provide more robust motivation for a 60 second latency.
\S~\ref{sec:prefilter} lists types of pre-filtered streams that could improve brokers' scientific productivity and estimates the resulting bandwidth reductions, and \S~\ref{sec:graceful} argues that when alerts are delayed, the scientifically optimal response would be to flag and distribute them as soon as possible (and not, e.g., send them directly to an archive).

%This draft contains content pertaining to the following open tickets in the DM-SST epic "Studies Around Alerts" (assignees):
%\begin{itemize}
%\item DM-19484, Estimate distribution of alert packet size based on scientific considerations (Bellm)
%\item DM-15654, Define policy for handling fields with $>$10K alerts (Bellm)
%\item DM-20296, Investigate the possibility of providing filtered streams (Graham)
%\item DM-20298, Revise the contents of the alert packets (Graham)
%\item DM-20299, Scientific use-cases requiring $<$5 minute access (Graham)
%\item DM-20300, Investigate options for expanding access to the alert stream (Bellm)
%\end{itemize}

\clearpage
% % % % % % % % % % % % % % % % % % % % % % % % % %
\section{Alert Packet Contents} \label{sec:packets}

In this section the scientific impacts of various options to reduce the size of alert packets (in order to, e.g., enable wider alert distribution, facilitate broker processing) are considered, as are potential additions to the alert packet contents which would improve the user's ability to classify and prioritize alert sources for follow-up.
The main conclusions of the following discussion are:
\begin{itemize}
\item {\bf Removing Histories (\S~\ref{sssec:packets_remove_hist}) ---} Removing the {\tt DIASource} histories from the alert packets that are transmitted to brokers that request this modification could be scientifically beneficial, but a list of all {\tt DIASource} identifiers associated with the {\tt DIAObject} at the time of alert generation must be added to those packets. 
\item {\bf Removing Stamps (\S~\ref{sssec:packets_remove_stamps}) ---} Removing the image stamps from all alert packets could be scientifically beneficial, but stamps should still be made during alert production and made publicly retrievable with a latency similar to that of alert distribution.
\item {\bf Distributing Multiple Packet Formats (\S~\ref{sssec:packets_remove_procon}) ---} From a science perspective, distributing non-identical alert packets would have no negative impact provided that full-sized packets do not become more prone to delays.
\item {\bf Adding Object Association Information (\S~\ref{ssec:packets_add}) ---} To better enable extragalactic transient science with brokers, two new {\tt DIAObject} catalog elements should be computed and included in the alert packets: (1) the {\tt objectId} for the three {\tt Object} catalog galaxies with the lowest separation distance (based on the galaxy's 2D luminosity profile) from the {\tt DIAObject}, and (2) the separation distances for those three {\tt Objects}.
\item {\bf Compressing Packets (\S~\ref{ssec:packets_compress}) ---} From a science perspective, compressing the alert packets has little scientific impact unless it imposes a large additional latency on alert distribution, which could be detrimental to alerts-related science.
\end{itemize}

\subsection{Removing Histories and/or Stamps}\label{ssec:packets_remove}

The size of a fully-loaded individual alert packet is estimated to be $\lesssim$82~KB, based on simulations of the planned content of the alerts as described in Section 3.5 of \citeds{LSE-163} (see also \citeds{DMTN-102}).
The two largest components of the alert packet which could be considered ``optional'' are the history and the postage stamps.
The history is the past $\sim$12 months of {\tt DIASource} records, previously released as alerts.
The postage stamps are, at minimum, 30$\times$30 pixels and contain flux, variance, and mask extensions for both the template and difference image, plus a header of metadata \citedsp{LSE-163}.
The history accounts for $\sim$27KB of the alert packet ($\sim$33\%) and the stamps contribute $\gtrsim$18~KB ($\sim$20\%).

Brokers that save all alerts, build their own a database of alerts contents, or have an automated interface with the LSST Prompt products database or Alerts database might not need the historical records to be included in the alert packets.
Brokers which do not use the image stamps -- or would need them for only a subset of the alerts and could feasibly query a stamps database -- might not need them included in the alerts packets.
Individual broker teams may indicate which information they require (or would like removed from the packets in their stream) during the broker proposal process \citedsp{LDM-612,LDM-682,LDM-723}.
During the first stages of the proposal process, and at the LSST Community Brokers Workshop\footnote{\url{ls.st/cbw}} in June 2019, there was some indication that at least a few brokers were interested in alerts without historical records and/or postage historical records.

\subsubsection{{\tt DIASource} Record History}\label{sssec:packets_remove_hist}

Including the {\tt DIASource} record history in every alert allows for the \emph{instant assessment} --- without the need to cross-match or query catalogs --- of changes in the objects's location, size, shape, or brightness, provides context for the latest detection (i.e., anomalous or typical), and enables a robust assessment of the likely physical nature of the object (e.g., light curve fits indicating supernova type become more reliable with more observations).

To remove the {\tt DIASource} record history from an alert would slightly inhibit the scientific assessment of transient/variable objects for brokers that build their own database of alert contents, due to the time required to cross-match with the database.
Removing the {\tt DIASource} record history would completely inhibit this kind of analysis for brokers that do not have such storage or processing capabilities.

There is also the issue that the association of {\tt DIASources} into {\tt DIAObjects} may change over time, for some objects.
For example, rare cases such as multiply-imaged strongly-lensed supernovae, or two unassociated transients/variables that are superimposed along the line of sight, might be blended/separated in poor/good seeing.
In such cases, the set of all associated {\tt DIASources} might be improved between one alert and the next.

{\bf Summary --} The {\tt DIASource} record histories should not be removed from all alert packets, but could be optionally removed at the request of a broker.
However, if the record histories are removed, a list of all {\tt DIASource} identifiers (the {\tt diaSourceId}, as in Section 3.3.1 of \citeds{LSE-163}) which are currently associated with the {\tt DIAObject} to which the alert-spawning {\tt DIASource} is associated should be provided (note that this list is \emph{not} currently part of the {\tt DIAObject} record).

\subsubsection{Forced Photometry History}\label{sssec:packets_remove_fp}

When forced photometry ({\tt DIAForcedSource} record history) is included in an alert packet, it is ``historical" in the sense that it is based on past images and not the image which spawned the alert.
However, it is not always redundant information that was previously released in an alert, like the {\tt DIASource} record history.

There are two main cases where forced photometry is performed, as described in \citeds{LSE-163}.
(1) For all new {\tt DIAObjects}, forced ``pre-covery'' photometry is done in the past 30 days of difference images within 24 hours, stored in the {\tt DIAForcedSource} catalog, and included in all future alert packets associated with that {\tt DIAObject}.
(2) For all {\tt DIAObjects} that are not detected\footnote{The detection threshold is a signal-to-noise ratio $\geq$ 5.} in a given difference image, but had a detection in the past $\sim$12 months, forced photometry is done on that difference image.
The result is stored in the {\tt DIAForcedSource} catalog, and included in all future alert packets associated with that {\tt DIAObject}.
The forced photometry is scientifically valuable because, for example, pre-covery forced photometry provides context for the first detection of a transient/variable, and low-SNR detections can help with photometric classification and prioritization for follow-up observations.

If a given {\tt DIAObject} is never again detected in a difference image, then the {\tt DIAForceSource} records are never included in an alert packet, but are available to individuals or brokers with access to the PPDB via the LSST Science Platform.
If a given {\tt DIAObject} is repeatedly detected in new difference images, then the same {\tt DIAForcedSource} records will be redundantly included in all future alerts (after the first one containing the pre-covery or non-detection forced photometry).
For the ``historical" {\tt DIAForcedSource} records, alert content redundancy could be avoided by only including previously undistributed {\tt DIAForcedSource} data in an alert packet.
To do this, the Prompt pipeline would have to either compare with the contents of the most recent alert and remove redundant data from the new alert, or flag the {\tt DIAForcedSource} data as ``released" and only add ``unreleased" forced photometry to a alert packet.
Both of these options would add processing times to the alert generation pipeline, which is undesirable.

{\bf Summary --} If the decision is made to remove the historical {\tt DIASource} records from alert packets, the {\tt DIAForcedSource} records of forced photometry should always be added to all alerts.
This will cause some redundancy in alert contents, but if not included there would be no public access to this scientifically valuable information.

\subsubsection{Postage Stamps}\label{sssec:packets_remove_stamps}

\textcolor{red}{MLG notes from DESC: not all alerts need them so it could be, e.g., for new DIASources only, non-stellar DIASources, etc. with the rest going to a server that does not have the 1 minute latency requirements.}

Including the postage stamps in the alert packet allows for the instant assessment of the {\tt DIASource} in the difference and template images, which has a variety of scientific applications.
For example, although real/bogus classification of the difference-image sources will be done prior to alert generation, some brokers may still wish to run additional algorithms on the images -- especially in the case of first detections.
Brokers might also employ algorithms to classify transients and variables based on the images instead of the photometry (e.g., \citealt{2019PASP..131j8006C}).
From the postage stamp images, scientifically useful information can be derived about the trails left by moving objects, or from the 2-dimensional luminosity distribution of spatially evolving objects such as comets, which might not be entirely captured by the {\tt DIASource} elements (see Table 1 in \citeds{LSE-163}).
The images also provide context for the {\tt DIASource}, such as host-galaxy morphology or field crowdedness, that can assist with follow-up observations.

If postage stamps were not created during alert production, the only option would be for users to wait until the images are made available (within 24 hours) in the LSP, which would significantly inhibit alerts-based science goals.
However, postage stamps are unlikely to be needed for most {\tt DIASources}, the majority of which --- 95\%, according to the breakdown in \citeds{DMTN-102} --- will be stars or asteroids: point sources for which context is not a significant part of their immediate evaluation.
An efficient compromise might be for the Prompt pipeline to create the postage stamps during alert generation, and then store them in a publicly accessible database (i.e., outside the LSP) where they can be obtained (via {\tt diaSourceId}) with a latency similar to the alert distribution timescale by anyone, from anywhere, at anytime.

The expected maximum download rate from this stamp database is estimated by assuming that all 5 community brokers download $\sim$500 stamp sets per visit: $\sim$2500 stamp sets every $\sim$35 seconds.
At 18~KB per stamp set, that's a data rate of 10.5 Mbps (for comparison, that's $\sim$5\% of one full alert stream).

{\bf Summary --} The postage stamps could be removed from all alert packets, but they should still be created during Prompt processing and stored in a publicly accessible database where they can be obtained with a latency similar to the alert distribution timescale.

\textcolor{red}{{\bf Could Postage Stamps Be Smaller? --}}
\citeds{LSE-163} describes the size of the postage stamp to be at least $6\arcsec \times 6\arcsec$ ($30\times30$ pixels), or large enough to encompass the entire footprint of the {\tt DIASource} (e.g., if trailed, extended). 
That is large enough to contain the angular diameter (30 kpc) of a Milky Way-like galaxy at a redshift of $z\approx0.5$, which will be a fairly typical supernova host galaxy in the LSST data set.
Since context is a science driver for producing stamps, it would not be desirable for them to be much smaller than $6\arcsec \times 6\arcsec$.

\subsubsection{Pros and Cons of Multiple Alert Formats}\label{sssec:packets_remove_procon}

The histories and/or stamps could be removed from all alert packets, or only at the request of a broker.
In the latter case, LSST might end up distributing four types of alert packet formats to its five brokers: (i) with histories and stamps, (ii) with histories but without stamps, (iii) without histories but with stamps, and (iv) without histories or stamps.

From a science perspective, the benefits to allowing brokers to customize their alert packet is that it potentially enables unique science to be done by brokers (e.g., very rapid custom stamp-analysis algorithms), and could enable LSST to support $>$5 brokers which might lead to more alerts-related science in general.

One drawback might arise during times of heavy load, such as many visits with $>$40,000 alerts (\S~\ref{sec:graceful}), \emph{if}, and only if, full-sized packet streams are more likely to be delayed.
Whether this becomes a possibility will depend on the technical aspects of how the multiple types of alerts are generated and transmitted, and also on the final decisions about latency (\S~\ref{sec:latency}) and delayed distribution (\S~\ref{sec:graceful}).
This may well never come to pass, but if it did, it could be unfair to brokers requiring full-size packets.
While some brokers are being designed for a wide variety of science goals, others are more specifically tailored.
If it is the specifically-tailored broker that requires full-sized alerts, then experiencing delayed transmission more frequently than other brokers could put some science goals at a disadvantage. 
Furthermore, the fact that not all alert packets are identical might cause some additional bookkeeping considerations for downstream brokers who ingest from multiple community brokers, but that should not be detrimental to their science goals.
The technical considerations of how many different alert packet formats it is feasible to create and transmit within the required latency time is beyond the scope of this document, in which we focus on science impacts.

{\bf Summary --} Offering multiple alert formats to brokers is scientifically beneficial provided that full-sized alerts are not predisposed to delay.

\subsection{Additional {\tt Object} Association Information}\label{ssec:packets_add}

\textcolor{red}{MLG notes from DESC: \\ (1) Check "Host Galaxy Identification for Supernova Surveys" \citet{2016AJ....152..154G}. \\ (2) Alex Gagliano is working on host association as a part of classification that does a PCA on galaxy properties, and found that color and morphology are the first two PCA components, and that the hosts of SNIa SNII etc. occupy different parts of the PC2 vs PC1 plane. His paper is in prep still. \\ (3) As MWV was trying to tell you, host association depends on the redshift of the object (i.e., you want to do it in absolute distance); obviously that won't be in the alert so we can't make it part of the "3 nearest", but you'll probably want to mention that specifically.}.

The contents of the alert packet as defined in \citeds{LSE-163} includes the following with respect to associations with {\tt Object} catalog from the most recent data release:
\begin{itemize}
\item {\tt nearbyObj} ({\tt unit64[6]}), the {\it "closest {\tt Objects} (3 stars and 3 galaxies) in Data Release database"}
\item {\tt nearbyObjDist} ({\tt float[6]}), the {\it "distances to {\tt nearbyObj}"} in arcseconds
\item {\tt nearbyObjLnP} ({\tt float[6]}), the {\it "natural log of the probability that the observed {\tt DIAObject} is the same as the nearby {\tt Object}"}
\end{itemize}
For the latter, there is a footnote that says {\it "This quantity will be computed by marginalizing over the product of position and proper motion error ellipses of the {\tt Object} and {\tt DIAObject}, assuming an appropriate prior"}.

The current definitions of {\tt nearbyObj}, {\tt nearbyObjDist}, and {\tt nearbyObjLnP} are not as useful as they could be for transients in host galaxies. 
For extragalactic transients, the three nearest galaxies are not always the three most likely host galaxies, and the distance in arcseconds matters less than a separation distance that accounts for the galaxy's luminosity profile.
Furthermore, the definition of {\tt nearbyObjLnP} is only appropriate for static variable point sources (stars): for transients in host galaxies, the observed {\tt DIAObject} will never be ``the same as the nearby {\tt Object}".

Statistically, the most likely host for a given transient is the galaxy which contributes the most optical flux at the transient's location.
This is usually estimated by calculating an offset distance from the nearby galaxies to the transient that is expressed in terms of the galaxy's spatial luminosity profile, and assuming the galaxy with the lowest offset distance is the host.
The following are several options for estimating which nearby galaxy is the most likely host of an extragalactic transient.

{\bf Effective Radius --} Calculate a separation distance that is the radial distance from the core of the galaxy to the location of the transient, divided by the effective radius of the galaxy (i.e., {\tt kronRad90} in the {\tt Objects} table; \citeds{LSE-163}).
The nearby galaxy with the lowest separation distance is the most likely host. This will account for the relative sizes of the potential host galaxies, but not their position angles.

{\bf Second Moment --} Calculate a separation distance, as in \citet{2006ApJ...648..868S}, based on the two-dimensional luminosity profile of the galaxy.
Where $x_{\rm trans},y_{\rm trans}$ is the location of the transient, and $x_{\rm gal},y_{\rm gal}$ is the center of the galaxy (the first moment; {\tt radec} in the {\tt Object} table), the separation distance is $R^2 = C_{xx} x_r^2 + C_{yy} y_r^2 + C_{xy} x_r y_r$, where $x_r = x_{\rm SN} - x_{\rm gal}$ and $y_r = y_{\rm SN} - y_{\rm gal}$.
The ellipse parameters $C_{xx}$, $C_{yy}$, and $C_{xy}$ can be calculated directly from the second moments of the galaxy luminosity profile ({\tt Ixx}, {\tt Iyy}, and {\tt Ixy} in the {\tt Object} table), e.g., as described in Section 10 of E. Bertin's Source Extractor manual\footnote{Version 2.3: \url{https://www.astromatic.net/pubsvn/software/sextractor/trunk/doc/sextractor.pdf}}.
The nearby galaxy with the lowest separation distance is the most likely host.
This option accounts for both the relative sizes and position angles of the potential host galaxies, and requires only a little more processing.

{\bf 2D Algorithms --} There are also other, more complicated methods for identifying the most likely host for a given transient.
For example, the nearby galaxy with the smallest fraction of light interior to an isophot through the transient's location, where the isophot shape is given more degrees of freedom and not constrained to concentric ellipticals as in the second moment method above.
Another example is to use an algorithm that provides deblended footprints for nearby extended objects, and can estimate the fraction of light in given pixel that should be attributed to each (e.g., the SCARLET deblender, \cite{2018A&C....24..129M}).
\textcolor{red}{\citeds{LSE-163} does mention that footprints will be generated for {\tt Objects}, but it doesn't look like they'd be stored in the {\tt Object} table; find out if these footprints would be appropriate/available to use here.}
The most likely host galaxy would be the one which contributes the most flux at the pixel location of a transient.

Generally, nearby galaxies with separation distances $>$3, or $>$99\% of the luminosity profile, are not assigned as hosts and the transient is considered "hostless".
While this cutoff has been appropriate for past samples of $\sim$hundreds of transients, it will not be appropriate for the LSST sample size, and no such cut should be applied when evaluating potential host galaxies for {\tt DIAObjects}.

All of the above host galaxy identification methods can also make use of priors if the transient type is known; for example, the established correlation between core-collapse supernovae and star formation.
However, such a robust level of host associations are beyond scope for the Prompt pipeline, and the main goal is to provide brokers with sufficient information to assess the potential host association, and decide whether it is useful to the classification and/or prioritization of alerts for follow-up.

{\bf Summary --} For the ten nearest {\tt Object} catalog galaxies, a separation distance should be calculated with respect to the transient location, preferably using the second moments of each galaxy's luminosity profile. Two new {\tt DIAObject} catalog elements should be added: {\tt nearbyPotHost}, containing the {\tt objectId} for the three galaxies with the lowest separation distances, and {\tt nearbyPotHostSepDist}, containing the separation distances. An analog for the existing element {\tt nearbyObjLnP}, representing the probability, is not necessary. This would add {\tt unit64[3]} and {\tt float[3]} to the {\tt DIAObject} catalog and to each alert, but this both tiny and worthwhile.

\subsection{Compressing the Alert Packets}\label{ssec:packets_compress}

The application of gzip compression could further reduce the size of a full alert to $\sim$65~KB (80\%; JIRA ticket DM-16280).
Naively, this seems like it would allow 1 more full stream to be transmitted to a broker, and thereby enable more science.
However, the time and computational resources required to compress the alert packets needs to be considered.
For example, gzip compression at 50 MB/s to compress $\sim$10000 alerts would take $\sim$10000 $\times$0.08 MB per alert $/$50 MB/s, or $\sim$16 seconds.
This is significant, considering the alert distribution latency requirement\reqparam{OTT1}\lsrreq{0101}\ossreq{0127}\dmreq{0004} is 60 seconds. 
Furthermore, compressing the alert packets forces brokers to then decompress the alerts on arrival, which would incur further delay.
Science goals requiring very low-latency alerts distribution might be negatively impacted by compression -- see \S~\ref{sec:latency} for a deeper discussion of the science drivers for low-latency alerts.

{\bf Summary --} Since packet compression could induce an additional latency, perhaps it should only be done if, e.g., an algorithm could provide a compression rate of $\lesssim$10 MB/s (requiring only a few seconds to compress a visit's worth of alerts), or if the alert distribution latency requirement is relaxed from 1 minute (see \S~\ref{sec:latency}).
Otherwise, compression would have no impact on alerts-related science goals and so this should be primarily a technically-driven decision.

\clearpage
% % % % % % % % % % % % % % % % % % % % % % % % % %
\section{Alert Distribution Latency} \label{sec:latency}

This section considers the scientific motivation for the 1 minute alert distribution latency, and the impact to alerts-related science if that latency was extended to, e.g., 5 minutes.
It is a requirement that the DMS be capable of supporting the distribution of at least 98\%\reqparam{OTT1}\reqparam{OTR1}\lsrreq{0101}\lsrreq{0025}\ossreq{0127}\dmreq{0004} of alerts for each visit within 60 seconds of the end of image readout.
However, if it is possible to relax this requirement without a negative impact on science then it might be possible for LSST to transmit alerts to $>$5 brokers (and/or repurpose human and computational resources dedicated to the 1 minute latency), which in turn might enable more alerts-related science.

Generally, for science goals that require alerts on timescales shorter than $\sim$5 minutes, this need stems from targets that are fast-evolving or short-lived (or both) and require follow-up on timescales shorter than $\sim$15--30 minutes.
Since most observing strategies for the WFD main survey do not revisit fields on timescales shorter than $\sim$15 minutes (strategies that do pair visits aim for closer to $\sim$22 minute gaps), brokers must be able to confidently identify such targets with single-epoch single-filter LSST photometry, or be using additional information (e.g., host galaxy characteristics, coincident alerts from other surveys).
For the purposes of this discussion, an optimistic outlook is assumed, and the difficulty in identifying targets with a single epoch of LSST photometry is not considered as a reason to downgrade the importance of a 1 minute alert latency.

After a consideration of the following science cases that are most likely to require rapid follow-up (summary below), it appears that most of them would be satisfied without risk if the latency was increased from 1 to 5 minutes. 
However, it is absolutely not unambiguously clear that a need for 1 minute alerts will not evolve out of the still emergent fields of fast radio bursts and gravitational wave events over the next couple of years.
\begin{itemize}
\item {\bf A TVS Survey (\S~\ref{ssec:latency_tvs})} found that only $<$10\% of respondents reported that requiring alerts within 1 minute was necessary for their science goals.
\item {\bf GW Follow-up (\S~\ref{ssec:latency_emgw}) ---} Although a 5 minute latency would suffice for kilonovae (days-long optical counterparts to neutron star mergers), some theoretical models predict optical emission within the first hour (e.g., jet break-out or spin-down energy from the merger). However, these GW scientific motivators for a 1 minute alert latency are moot if the time between GW event and LSST ToO follow-up remains at tens of minutes or more.
\item {\bf Gamma-Ray Bursts (\S~\ref{ssec:latency_grb}) ---} \textcolor{red}{Tentatively,} alerts for the coincident optical counterparts of short GRBs would be most scientifically useful if distributed within 1 minute, but the predicted occurrence rate is $\lesssim$20 over the 10-year LSST survey.
\item {\bf Fast Radio Bursts (\S~\ref{ssec:latency_frb}) ---} \textcolor{red}{Tentatively,} some models predict that a short, bright burst of optical emission from the physical mechanism powering FRBs could be detected by LSST 5 to 15 minutes prior to the radio signal.
If such optical events occur with significant frequency, are bright and consistently detectable, and can be identified with a single-filter detection, and if radio telescopes can respond within minutes, then LSST is the most promising future facility for this science, which requires a 1 minute alert distribution latency.
\item {\bf Young Supernovae (\S~\ref{ssec:latency_ysne}) ---} The science goals require follow-up in an hour or longer, and a 5 minute latency would not have a negatively impact.
\item {\bf Solar System Objects (\S~\ref{ssec:latency_sso}) ---} \textcolor{red}{TBD.}
\item {\bf Stellar Variables (\S~\ref{ssec:latency_stars}) ---} \textcolor{red}{TBD.}
\end{itemize}

@@

\clearpage
% % % % % % % % % % % % % % % % % % % % % % % % % %
\section{Pre-Filtered Alert Streams} \label{sec:prefilter}

This section explores the scientific motivation and impact of offering brokers pre-filtered alert streams. 
Pre-filtering could reduce the overall bandwidth of alert distribution and potentially allow streams to be transmitted to more brokers, thereby enabling more alerts-related science.
Providing pre-filtered streams could also lessen the load and resources used by the brokers, enabling more alerts-related science by allowing brokers to spend a smaller fraction of their budget computational resources.

Depending on the science goals of a given broker and/or their users, a broker might find it useful to only receive alerts for {\tt DIASources} which:
\begin{itemize}
\item are likely to have a certain physical origin, e.g., stellar variable, moving objects
\item could potentially be observed by the follow-up resources its users have access to
\item have at least one prior detection by LSST (because, e.g., their algorithms all require photometric confirmation or a change in brightness)
\item originate in images obtained as part of a special program such as the deep drilling fields or target-of-opportunity imaging surveys
\item were not obtained during poor weather conditions (e.g., an image quality or sky background violating some limit)
\end{itemize}

For the above science drivers, the types of pre-filters that are likely to be scientifically useful -- some of which have been suggested by brokers already -- include:
\begin{itemize}
\item {\bf Apparent Magnitude:} only {\tt DIASources} brighter than a specified apparent magnitude (for each of the LSST filters {\it ugrizy}) are transmitted. 
\item {\bf Region of Sky:} only {\tt DIASources} in a given sky area (or areas; defined by, e.g., limits in right ascension or declination), are transmitted. 
\item \texttt{\bf DIAObject} {\bf History:} only {\tt DIASources} associated with {\tt DIAObjects} that have at least $N$ previous detections are transmitted, where $N$ can be $\geq1$. 
\item \texttt{\bf SSObject} {\bf Association:} only {\tt DIASources} that are associated with a {\tt SSObject} are transmitted. (Potentially including new, unassociated {\tt DIASources}, because they might also be moving objects).
\item \texttt{\bf Object} {\bf Association:} only {\tt DIASources} associated with {\tt DIAObjects} that are, in turn, associated with a data release {\tt Object} that meets some specification are transmitted (e.g., associated with a stellar point source).
\item {\bf Observation Metadata:} only alerts from images obtained with a given set of metadata, e.g., a special observing program or meeting sky condition thresholds, are transmitted.
\end{itemize}

All of these potential science-driven pre-filters are based on the information contained in the alert itself, and do not impose any additional processing on the Prompt pipeline.
They all also have the potential to considerably reduce the number of transmitted alerts.
Due to the exponential relationship between the number of variable stars and their variability amplitude, and also that of volume and distance modulus, an apparent magnitude limit $\sim1$ mag brighter than the nominal $5{\sigma}$ {\tt DIASource} detection limit could reduce the alert stream data rate by $\gtrsim50$\%.
Limiting to alerts that are associated with past detections or certain types of objects could similarly reduce the alert stream data rate.
These types of pre-filers would reduce the number of alerts per visit for all visits -- the average bandwidth -- which might enable alert distribution to extend to more brokers and thus enable more alerts-related LSST science.
Pre-filters based on their sky location or observational metadata would probably either deliver all or none of the alerts for a given visit, causing fluctuations in the bandwidth but not necessarily in a way that would allow for additional streams to be transmitted.

% % % % % % % % % % % % % % % % % % % % % % % % % %
\subsection{LSST Alert Filtering Service}\label{ssec:prefilter_lafs}

As described in \citeds{LSE-163} the LSST will provide an alert filtering service with limited capacity, by which individuals with LSST data rights and access may receive alerts via pre-defined filters and/or create and apply their own filters to the stream \citedsp{LPM-17,LSE-61}.
It is a requirement that the LSST alert filtering service be able to support\reqparam{numBrokerUsers}\dmreq{0343} at least 100 simultaneous users, and capable of returning 20\reqparam{numBrokerAlerts}\dmreq{0343} full-sized alerts per visit per user.
User-generated filters may be comprised of, e.g., a series of {\tt if} statements applied to the alert packet elements.
There are no requirements on the latency of filter execution, which may depend on the complexity of the filter and available resources.

>>>>

% To add a short-form title:
% \title[Short title]{Title}
\title[Alerts Latency]{Science Drivers for LSST Alert Distribution Latency}

@@

\author{%
M.~L.~Graham, with contributions from the Transients and Variable Stars Science Collaboration
}

@@

\setDocAbstract{%
\textcolor{red}{\bf EXTREME DRAFT STATE. DO NOT CITE. DO NOT READ. LOOK AWAY.}\\
A discussion of the science drivers regarding the alert packet distribution latency.
}

% Change history defined here.
% Order: oldest first.
% Fields: VERSION, DATE, DESCRIPTION, OWNER NAME.
% See LPM-51 for version number policy.
\setDocChangeRecord{%
  \addtohist{0}{2019-11-14}{Inception.}{Melissa Graham}
  \addtohist{0.1}{2020-01-09}{Early draft.}{Melissa Graham}
}

@@

This document considers the science motivation and impacts of the alert distribution timescale, also called the latency.
Alert packets will play a major role in the scientific impact of LSST, in part because they are the only LSST data product that is both world public (exempt from any proprietary period) and publicly accessible, as they are distributed to community brokers and not solely available via the LSST Science Platform (LSP).
In comparison, although the contents of the Prompt products database (PPDB) from which alerts are generated are also world public, access to the PPDB is restricted to individuals with data rights \citedsp{LDO-013} who have authenticated accounts in the LSP.

It is a requirement\reqparam{numStreams}\dmreq{0391} that the DMS be capable of supporting the transmission of at least 5 full alert streams within 60 seconds of image readout --- more specifically, that the data management system (DMS) be capable of supporting the distribution of at least 98\%\reqparam{OTT1}\reqparam{OTR1}\lsrreq{0101}\lsrreq{0025}\ossreq{0127}\dmreq{0004} of alerts for each visit within 60 seconds of the end of image readout.
The 5 full streams aspect is based in part on estimates of the alert stream data rate and the bandwidth allocated to alert distribution in the LSST data facility \citedsp{DMTN-102}.
The 60 second aspect is scientifically motivated by a desire to explore the parameter space for very short timescale transients and variables which might last only a few minutes \citedsp{LPM-17}.

Increasing the alert latency could enable the stream to be delivered to more brokers, thereby increasing the amount of alerts-related science done by the community.
Or, it might allow human and computational resources currently dedicated to building a DMS that can deliver a 1 minute latency to be repurposed. 
In case it ever becomes technically necessary to change the alert latency, this document collects the science drivers for short timescale alert distribution.
\textbf{Time-domain astronomy remains a rapidly evolving field and this should be considered a living, in-progress document}.

\subsection{Summary}

This document reports that most alerts-related science goals would be satisfied with a $\sim$5 minute alert timescale, but that the quickly evolving fields of fast radio bursts and gravitational wave events may soon provide more robust motivation for a 60 second latency.

%This draft contains content pertaining to the following open tickets in the DM-SST epic "Studies Around Alerts" (assignees):
%\begin{itemize}
%\item DM-20299, Scientific use-cases requiring $<$5 minute access (Graham)
%\end{itemize}

\clearpage
% % % % % % % % % % % % % % % % % % % % % % % % % %
\section{Science Drivers for Alert Latency} \label{sec:latency}

Generally, for science goals that require alerts on timescales shorter than $\sim$5 minutes, their need stems from targets that are fast-evolving or short-lived (or both) and require follow-up on timescales shorter than $\sim$15--30 minutes.
Since most observing strategies for the WFD main survey do not revisit fields on timescales shorter than $\sim$15 minutes (strategies that do pair visits aim for closer to $\sim$22 minute gaps), brokers must be able to confidently identify such targets with single-epoch single-filter LSST photometry, or be using additional information (e.g., host galaxy characteristics, coincident alerts from other surveys).
For the purposes of this discussion, an optimistic outlook is assumed, and the difficulty in identifying targets with a single epoch of LSST photometry is not considered as a reason to downgrade the importance of a 1 minute alert latency.

\begin{itemize}
\item {\bf A TVS Survey (\S~\ref{ssec:latency_tvs})} found that only $<$10\% of respondents reported that requiring alerts within 1 minute was necessary for their science goals.
\item {\bf GW Follow-up (\S~\ref{ssec:latency_emgw}) ---} Although a 5 minute latency would suffice for kilonovae (days-long optical counterparts to neutron star mergers), some theoretical models predict optical emission within the first hour (e.g., jet break-out or spin-down energy from the merger). However, these GW scientific motivators for a 1 minute alert latency are moot if the time between GW event and LSST ToO follow-up remains at tens of minutes or more.
\item {\bf Gamma-Ray Bursts (\S~\ref{ssec:latency_grb}) ---} \textcolor{red}{Tentatively,} alerts for the coincident optical counterparts of short GRBs would be most scientifically useful if distributed within 1 minute, but the predicted occurrence rate is $\lesssim$20 over the 10-year LSST survey.
\item {\bf Fast Radio Bursts (\S~\ref{ssec:latency_frb}) ---} \textcolor{red}{Tentatively,} some models predict that a short, bright burst of optical emission from the physical mechanism powering FRBs could be detected by LSST 5 to 15 minutes prior to the radio signal.
If such optical events occur with significant frequency, are bright and consistently detectable, and can be identified with a single-filter detection, and if radio telescopes can respond within minutes, then LSST is the most promising future facility for this science, which requires a 1 minute alert distribution latency.
\item {\bf Young Supernovae (\S~\ref{ssec:latency_ysne}) ---} The science goals require follow-up in an hour or longer, and a 5 minute latency would not have a negatively impact.
\item {\bf Solar System Objects (\S~\ref{ssec:latency_sso}) ---} \textcolor{red}{TBD.}
\item {\bf Stellar Variables (\S~\ref{ssec:latency_stars}) ---} \textcolor{red}{TBD.}
\end{itemize}

@@
